---
title: "patchclampplotteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{patchclampplotteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(reactable)
library(dplyr)
```

# How to use patchclampplotteR

This package is designed to help you import, analyze, and plot your patch clamp
data efficiently. This vignette will walk you through a complete example of
using `patchclampplotteR` to go from raw data to publication-quality plots!

```{r setup}
library(patchclampplotteR)
```

## About the data

This dataset consists of whole-cell patch clamp recordings of neurons within the
dorsomedial hypothalamus (DMH), a brain region critical for appetite regulation,
stress responses, thermoregulation, and other processes. I recorded evoked
excitatory post-synaptic currents for five minutes under baseline conditions,
then added 500 nM insulin to the perfusion solution, and I continued recording
for 25 minutes. For more detail, please see the article "About the Data"

## Import Data

My goal is to determine if insulin affects evoked current amplitude in DMH neurons.

First, I must import a csv file containing information about factors such as the
animal's age and sex, the cell ID number, and other details.

> Note: Since this vignette uses a dataset included within an R package, the following code requires the function `import_ext_data()` to properly locate the csv file. For anyone using the package, you can likely just write the filename directly within `import_cell_characteristics_df()`.

```{r}
cell_characteristics <- import_cell_characteristics_df(import_ext_data("sample_cell_characteristics.csv"))

reactable(cell_characteristics, defaultPageSize = 5)
```


