---
title: "patchclampplotteR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{patchclampplotteR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knit-options-libraries, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(reactable)
library(dplyr)
```

# How to use patchclampplotteR

`patchclampplotteR` will help you analyze and plot your patch clamp data
efficiently. This vignette will walk you through the complete process of
transforming raw data into publication-quality plots!

## Install and load package

You can install the development version of patchclampplotteR from
[GitHub](https://github.com/). Only do this once per computer, or if there's a major update.

```{r install-package-demo, eval=F, warning=FALSE, message=FALSE}
pak::pak("christelinda-laureijs/patchclampplotteR")
```

And then load the package each time you want to use it:

```{r load-library}
library(patchclampplotteR)
```

## About the data

This dataset consists of whole-cell patch clamp recordings of neurons within the
dorsomedial hypothalamus (DMH), a brain region critical for appetite regulation,
stress responses and other processes. I recorded evoked excitatory post-synaptic
currents for five minutes under baseline conditions, then added 500 nM insulin
to the perfusion solution, and I continued recording for 25 minutes.

> My goal is to determine if insulin affects evoked current amplitude in DMH
neurons.

## Import raw .csv files

First, I must import a `.csv` file containing information about factors such as
the animal's age and sex, the cell ID number, and other details. Please see the
**Required columns** section of the documentation for
`import_cell_characteristics_df()` for explanations of the required columns and
what they include.

> Note: Since this vignette is included within an R package, the following code
requires the function `import_ext_data()` to properly locate the `.csv` file in
the package folder. This won't be required when you are using the package within
your own project folder.
>
> You can just write the path to the filename
directly within `import_cell_characteristics_df()`. For example, you should
write `import_cell_characteristics_df("Data/cell_info.csv")` to use
`cell_info.csv` located within the `Data/` folder.

```{r import-cell-characteristics}
cell_characteristics <- import_cell_characteristics_df(import_ext_data("sample_cell_characteristics.csv"))

reactable(cell_characteristics)
```

Next, I will import the raw evoked current data that has been copied over from
Clampfit. This is a `.csv` file containing four columns: `Letter`, `ID`, `P1`
and `P2`:

* `letter`: A unique identifier for a single recording, which allows you to link evoked current data, spontaneous current data, action potential, data, and information on cell characteristics.

* `ID`: The name of the .abf filename used to obtain the data, which is useful for verifying the recordings and cross-referencing to your lab book.

* `P1`: The amplitude of the first evoked current (pA).

* `P2`: The amplitude of the second evoked current (pA).

> Don't worry too much about capitalizing the column names or not.
`add_new_cells()` will automatically convert all column names to lowercase for
consistency across functions. Capitalized letters will be retained for columns
like `ID`, `X`, `Y`, `P1`, and `P2`.

```{r sample-eEPSC-data}
sample_eEPSC_data <- read.csv(import_ext_data("sample_new_eEPSC_data.csv"))

reactable(sample_eEPSC_data)
```

## Add new cells

The next step is to merge the raw evoked current data with the cell
characteristics data. `add_new_cells()` will merge these two datasets, using
`letter` as the common column. This function requires three `.csv` files:

* The new raw data
* The cell characteristics
* An existing `.csv` with raw data that has been previously imported. As your
project goes on, you will eventually be appending new data onto your existing
datasheet, but if you are starting completely fresh, use a blank `.csv` file
containing just the required column names.

```{r blank-sheet-example}
blank_csv <- read.csv(import_ext_data("empty_raw_eEPSC_datasheet.csv"))

reactable(blank_csv)
```

Here is an example of what a full dataset can look like. In the following
example, I've just added new data for one cell (`letter = AV`).

```{r add-new-cells}
raw_eEPSC_data <- add_new_cells(
  new_raw_data_csv = import_ext_data("sample_new_eEPSC_data.csv"),
  cell_characteristics_csv = import_ext_data("sample_cell_characteristics.csv"),
  old_raw_data_csv = import_ext_data("sample_eEPSC_data.csv"),
  current_type = "eEPSC",
  write_new_csv = "no",
  new_file_name = "20241118-Raw-eEPSC-Data.csv"
)
```

**Output messages**

`add_new_cells()` pruduces several warnings and messages. One warning lets you
know you know that the column names have been renamed to lowercase. This is to
avoid case-sensitive issues from appearing in later functions.

The first message generated with `add_new_cells()` indicate that the
`sample_cell_characteristics.csv` and `sample_new_eEPSC_data.csv` have the same
cells. This is useful to catch if you forget to add the cell characteristics for
the new data.

The second message indicates that all letters in the new data are new relative
to the existing dataset. This ensures that you don't accidentally paste in the
same data twice, resulting in duplicated data.

The final message prints a list of the letters that have been added to the
dataset. In this case, this is `AV`. It is a good way to confirm that you've
added the letters you were planning to add.

You can also ask R to produce a list of all of the unique letters in the dataset. This won't catch duplicates, but it can help you identify if a letter is completely missing from the dataset. See, `AV` is now included.

```{r view-unique-letters}
unique(raw_eEPSC_data$letter)
```


## Analyze data

### Step 1: Normalize currents

The first step is to normalize the current amplitudes within each recording
relative to the average current amplitude during the baseline period. This makes
it easier to compare across cells that have a wide range of starting amplitudes,
since all baseline values will be converted to (roughly) 100%.

> Note how I set the minimum and maximum time values. This will limit the data
to values between 0 min and 25 minutes.
>
> I set the `interval_length` to 5 because I wanted to divide my data into
5-minute intervals for later statistical analyses.
>
> The baseline period (`baseline_length`) lasted 5 minutes.
> Clampfit recorded the current amplitude as negative values, so I set
`negative_transform_currents` to "yes" which will flip the current amplitudes to
positive values.

```{r normalize-data}
raw_eEPSC_df <- make_normalized_EPSC_data(
  filename = import_ext_data("sample_eEPSC_data.csv"),
  current_type = "eEPSC",
  min_time_value = 0,
  max_time_value = 25,
  interval_length = 5,
  baseline_length = 5,
  negative_transform_currents = "yes"
)
```

This is the sample raw evoked current dataset included in the package. To reduce
the vignette size, I am printing just the first 20 rows. The full dataset
contains > 5680 rows!)

```{r view-sample-raw-data}
# To reduce the vignette size, I am printing just the first 20 rows. The 
head(raw_eEPSC_df, n = 20) %>% 
  reactable()
```

`make_normalized_EPSC_data()` will retain the cell characteristics and `P1` and
`P2` values from before. However, you will notice some changes.

If you set `negative_transform` to "yes", `P1` and `P2` will be multiplied by
`-1`. This is to "flip" current amplitude data that was recorded as negative
values in Clampfit. Since the data are evoked current data (`current_type =
"eEPSC"`), some new columns are added. They are:

* `PPR`: The paired-pulse ratio, which is the amplitude of the second evoked current divided by the first evoked current (`PPR = P2/P1`).

* `interval`: The interval that the data belongs to. I set the `interval_length` to 5, which means the data will be divided into 5-minute intervals. The intervals will have names like "t0to5", "t5to10", and so on up until the maximum interval.

* `baseline_range`: You probably won't interact with this much, but this is just a column stating "TRUE" if the time is within the baseline period, or "FALSE" if the time is outside of this range. This is required for the normalization function to identify which values are outside of the baseline (and should be transformed).

* `baseline_mean`: This is one number that represents the average evoked current amplitude during the baseline period. This value is different for each recording.

* `P1_transformed`: The first evoked current amplitude, normalized relative to the mean baseline amplitude. For example, if the mean baseline amplitude is **80 pA**and a `P1` value is **40 pA**, `P1_transformed` will be **50%**.

* `P2_transformed`: The second evoked current amplitude, normalized relative to the mean baseline amplitude of the first evoked current.

#### Plot raw data

Let's see what the raw values look like over time!

`plot_raw_current_data()` will generate a scatterplot of evoked current
amplitude (pA) over time (min) for all cells within the treatment and category
that you specify. Behind the scenes, this really runs a loop over each letter,
generating a ggplot object for each recording.

Please see the documentation for `plot_raw_current_data()` to learn about the
arguments in more detail.

```{r make-raw-plots}
raw_eEPSC_control_plots <- plot_raw_current_data(
  data = raw_eEPSC_df,
  plot_treatment = "Control",
  plot_category = 2,
  current_type = "eEPSC",
  parameter = "P1",
  pruned = "no",
  hormone_added = "Insulin",
  hormone_or_HFS_start_time = 5,
  theme_options = sample_theme_options,
  treatment_colour_theme = sample_treatment_names_and_colours
)
```

`plot_raw_current_data()` will return a list of ggplot objects. If you want to
observe just one specific plot, you can select it by letter.

```{r view-one-raw-plot, fig.cap = "Insulin decreased evoked current amplitudes in this cell!"}
raw_eEPSC_control_plots$L
```

### Step 2: Prune data

It is often useful to summarize the data per minute. If you are familiar with
GraphPad Prism's "prune rows" function, `make_pruned_EPSC_data()` will perform
the same function.

> In this vignette, I'll use the example of pruning data per minute because this
is what is typically used in the Crosby lab. You can change this value by
changing the `interval_length` to something other than `1`.

```{r prune-data}
pruned_eEPSC_df <- make_pruned_EPSC_data(
  data = raw_eEPSC_df,
  current_type = "eEPSC",
  min_time_value = 0,
  max_time_value = 25,
  baseline_length = 5,
  interval_length = 1
)
```

This function will return a list of three dataframes. To access each list, type the object name, followed by a dollar sign. For example, write `pruned_eEPSC_df$individual_cells` to access the first dataframe in the list.

The three dataframes are:

*`$individual_cells`: This dataframe has the same structure as the raw evoked current data, except the data have been pruned per minute. New columns include `mean_P1` and `sd_P1`, and there are some other columns for variance analysis (please see the documentation for `make_pruned_EPSC_data()` for more details).

*`$for_table`: This dataframe has only two columns: `letter` and `P1_transformed` where the pruned `P1` values have been collapsed into a list. This is used to create a sparkline in `make_interactive_summary_table()`.

*`$all_cells`: This dataframe contains data that have been grouped by treatment and sex. In this dataframe, the data have been summarized and collapsed into one datapoint per minute for all cells per minute for a specific sex. This is useful for creating summary plots for publication (e.g. `plot_summary_current_data()`) and for future statistical testing to compare groups.

#### Plot pruned data

You can use the same `plot_raw_current_data()` to plot the pruned data. You will
need to make changes to the following arguments:

* `data`: Refer to the third element of the list produced from `make_pruned_EPSC_data()`. This is `$individual_cells`.
* `parameter`: Change this to "mean_P1".
* `pruned`: Change this to "yes"

```{r plot-pruned-current-data}
raw_eEPSC_control_plots_pruned <- plot_raw_current_data(
  data = sample_pruned_eEPSC_df$individual_cells,
  plot_treatment = "Control",
  plot_category = 2,
  current_type = "eEPSC",
  parameter = "mean_P1",
  pruned = "yes",
  hormone_added = "Insulin",
  hormone_or_HFS_start_time = 5,
  theme_options = sample_theme_options,
  treatment_colour_theme = sample_treatment_names_and_colours
)
```

```{r plot-pruned-cell-L, fig.cap = "See how this is the same as the raw data plot, except for it is pruned per minute?"}
raw_eEPSC_control_plots_pruned$L
```

The pruned data from all cells within a specific treatment and sex (`$all_cells`) will enable you to make a summary plot using `plot_summary_current_data()`.

> Notice how `data` is now `sample_pruned_eEPSC_df$all_cells`, and `parameter` is "amplitude". There are lots of customization opportunities here, including adding a representative trace as a .png overlay, which you can read about in the documentation for this function.

```{r plot-summary-data}
plot_summary_current_data(
  plot_category = 2,
  plot_treatment = "Control",
  data = sample_pruned_eEPSC_df$all_cells,
  current_type = "eEPSC",
  parameter = "amplitude",
  include_representative_trace = "no",
  signif_stars = "yes",
  t_test_df = sample_eEPSC_t_test_df,
  hormone_added = "Insulin",
  large_axis_text = "no",
  shade_intervals = "no",
  hormone_or_HFS_start_time = 5,
  treatment_colour_theme = sample_treatment_names_and_colours,
  theme_options = sample_theme_options
)
```


### Step 3: Summarize data

The next step is to group the data by treatment by sex and obtain summary data.
This dataset groups the data into intervals and generates summary statistics
(like mean and standard error) for each point. The interval length was already
specified during the `make_normalized_EPSC_data()` function from earlier.

```{r make-summary-data}
summary_eEPSC_df <- make_summary_EPSC_data(
  data = sample_raw_eEPSC_df,
  current_type = "eEPSC",
  save_output_as_RDS = "no"
)

reactable(summary_eEPSC_df)
```


## Other analyses

### Paired-pulse ratio

#### Create PPR dataset

The function `make_PPR_data()` is actually just a filtering function that will limit the raw evoked current data to two specific intervals. These represent the "before" (`baseline_interval`) and "after" (`post_hormone_interval`) states. You can also choose to limit the PPR values to a certain range to exclude outliers.

```{r make-ppr-data}
PPR_df <- make_PPR_data(
  data = raw_eEPSC_df,
  include_all_treatments = "yes",
  list_of_treatments = NULL,
  PPR_min = 0,
  PPR_max = 5,
  baseline_interval = "t0to5",
  post_hormone_interval = "t20to25",
  treatment_colour_theme = sample_treatment_names_and_colours
)

head(PPR_df, n = 10) %>% 
  reactable()

```


#### Plot PPR data

For a specific treatment:

```{r plot-one-ppr-treatment}
plot_PPR_data_one_treatment(
  data = PPR_df,
  plot_treatment = "Control",
  plot_category = 2,
  baseline_label = "Baseline",
  post_hormone_label = "Insulin",
  large_axis_text = "no",
  treatment_colour_theme = sample_treatment_names_and_colours,
  theme_options = sample_theme_options,
  save_plot_png = "no"
)
```

For multiple treatments:

```{r plot-multiple-ppr-treatments}
plot_PPR_data_multiple_treatments(
  data = PPR_df,
  include_all_treatments = "yes",
  plot_category = 2,
  baseline_label = "B",
  post_hormone_label = "I",
  theme_options = sample_theme_options,
  treatment_colour_theme = sample_treatment_names_and_colours
)
```

### Variance analysis

#### Create variance dataset

```{r make-variance-dataset}
variance_df <- make_variance_data(
  data = summary_eEPSC_df,
  df_category = 2,
  include_all_treatments = "yes",
  list_of_treatments = NULL,
  baseline_interval = "t0to5",
  post_hormone_interval = "t20to25",
  treatment_colour_theme = sample_treatment_names_and_colours,
  save_output_as_RDS = "no"
)

reactable(variance_df)
```


#### Plot variance comparisons

You can create plots comparing the inverse coefficient of variation squared, and the variance-to-mean ratio.

```{r}
cv_comparison_control_plot <- plot_variance_comparison_data(
  data = variance_df,
  plot_category = 2,
  plot_treatment = "Control",
  variance_measure = "cv",
  baseline_interval = "t0to5",
  post_hormone_interval = "t20to25",
  post_hormone_label = "Insulin",
  large_axis_text = "no",
  treatment_colour_theme = sample_treatment_names_and_colours,
  theme_options = sample_theme_options
)

vmr_comparison_control_plot <- plot_variance_comparison_data(
  data = variance_df,
  plot_category = 2,
  plot_treatment = "Control",
  variance_measure = "VMR",
  baseline_interval = "t0to5",
  post_hormone_interval = "t20to25",
  post_hormone_label = "Insulin",
  large_axis_text = "no",
  treatment_colour_theme = sample_treatment_names_and_colours,
  theme_options = sample_theme_options
)

cv_comparison_control_plot
vmr_comparison_control_plot
```

### Baseline comparisons


```{r}
plot_baseline_data(
  data = summary_eEPSC_df,
  current_type = "eEPSC",
  parameter = "raw_amplitude",
  include_all_treatments = "yes",
  list_of_treatments = NULL,
  baseline_interval = "t0to5",
  large_axis_text = "no",
  plot_width = 8,
  treatment_colour_theme = sample_treatment_names_and_colours,
  theme_options = sample_theme_options,
  save_plot_png = "no"
)
```

