---
title: "FAQ"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(ggplot2)
library(patchclampplotteR)
library(here)
library(reactable)
library(extrafont)
font_import()
```

# Plotting FAQ

This article contains answers to common questions about plots and customizing
the ggplot output of functions like `plot_raw_current_data()`.
This is the plot we will use throughout the FAQ:

```{r}
cars_plot <- ggplot(cars, aes(x = speed, y = dist)) +
  geom_point() +
  labs(x = "Speed (mph)", y = "Distance (ft)") +
  patchclampplotteR_theme()

cars_plot
```


## How do I save my plots outside of R?

Most plots have an argument called `save_png`, which you can set to "yes" or
"no" (the default is "no" to cut down on run time). If `save_png` is set to
"yes", this will save the plot as a `.png` using `ggsave()` and export it to a
folder. The subfolder will vary depending on the plot type. For example, all
plots generated using `plot_PPR_data_one_treatment()` will be exported to
"Figures/Evoked-currents/PPR" relative to your project directory. If you want
further control over the export options, you can also use `ggsave()` to manually
save a ggplot object. 

```{r, eval=F}
ggsave(
  cars_plot,
  path = here("Figures/Raw-plots"),
  file = "cars-plot.png",
  width = 7,
  height = 5,
  units = "in",
  dpi = 300
)
```


## How do I change the plot text sizes? 

A quick fix for making the text bigger is the `large_axis_text` argument. Set
`large_axis_text` to "yes" to add a ggplot theme layer which increases and
adjusts the plot to be ideal for posters or presentations. If you want more
control over the font sizes,  you can also apply a new `theme()` layer and
specify the text elements manually. For example:

```{r}
cars_plot +
  theme(text = element_text(size = 25, color = "darkmagenta"))
```

## How do I change the plot font family?

This can be tricky and it will vary depending on your operating system. I would
recommend using the `extrafont` package. You will need to run the code below the
first time you do this. Use `fonts()` to see the list of fonts available to you in R.

> **Warning!** `font_import()` will take a long time to
run, especially if you have a lot of fonts on your computer. Luckily, you only
need to do this once per system, or after you install a new font.

```{r, eval=F, results=F}
library(extrafont)
font_import()

fonts()
```

If you already have the fonts installed in R, you just need to run
`library(extrafont)` at the start of your document to have the fonts available to use. Here is a silly example:

```{r}
cars_plot +
  theme(text = element_text(size = 25, color = "darkmagenta", family="Jokerman"))
```


To apply these theme changes globally (i.e. all plots in your file), it will be much more efficient to modify and define a theme. See below!

## How do I modify the ggplot theme?

You might wonder things like, "How can I override the default theme that came with this package?" or "How can I make all my plots have the same theme/fonts, etc.?". You can fix this efficiently by defining a custom ggplot theme and making it the default theme for a document.

At the top of your RMarkdown file, choose a theme like `theme_classic()` or the `patchclampplotteR_theme()` that came with this package. Use this theme as a base, and add new theme elements to replace just the components you want to change (see [the documentation on theme modification in ggplot2](https://ggplot2.tidyverse.org/reference/theme.html). The new theme that you define will inherit all components of the base theme, but replace just the elements that you specified.

```{r}
my_new_theme <- patchclampplotteR_theme() %+replace%
  theme(text = element_text(size = 25, color = "darkmagenta", family="Jokerman"))
```

```{r}
cars_plot + my_new_theme
```


Once you are happy with your theme, define it as the default theme using `theme_set()`.

```{r}
theme_set(theme_gray())
```


Now, all of the plots in your document will have your specific theme, font families, etc. automatically applied. This will overwrite the default `patchclampplotteR_theme()` as well.

```{r}
plot_raw_current_data(
  data = sample_raw_eEPSC_df,
  plot_treatment = "Control",
  plot_category = 2,
  current_type = "eEPSC",
  y_variable = "P1",
  pruned = "no",
  hormone_added = "Insulin",
  hormone_or_HFS_start_time = 5,
  theme_options = sample_theme_options,
  treatment_colour_theme = sample_treatment_names_and_colours,
  save_plot_png = "no"
)
```


# Theme FAQ

## How do I insert my own colours and treatment names?

You will define your treatment names and colours once at the start, then refer to it using the `treatment_colour_theme` in all plotting functions.

This package comes pre-loaded with `sample_treatment_names_and_colours`:

```{r}
reactable(sample_treatment_names_and_colours)
```

First, check out how many treatment groups you have using unique(raw_eEPSC_df$treatment).

```{r}
unique(sample_raw_eEPSC_df$treatment)
```

Next, modify this code to set up your own dataframe with your own hormone names and colours.

```{r}
my_theme_colours <- data.frame(
  treatment = c("Control", "HNMPA", "PPP", "PPP_and_HNMPA"),
  display_names = c("Control", "HNMPA", "PPP", "PPP\n&\nHNMPA"),
  colours = c("#f07e05", "#f50599", "#70008c", "#DCE319"),
  very_pale_colours = c("#fabb78", "#fa98d5", "#ce90de", "yellow")
)
```

Every time a plot contains the argument `treatment_colour_theme`, refer to your custom dataframe.

```{r}
plot_PPR_data_multiple_treatments(
  data = sample_PPR_df,
  include_all_treatments = "yes",
  plot_category = 2,
  baseline_label = "B",
  post_hormone_label = "I",
  test_type = "t.test",
  theme_options = sample_theme_options,
  treatment_colour_theme = my_theme_colours
)
```

## How do I modify the theme_options?

This dataset contains theme options that will affect all plots. These include values for line thickness and the point shapes for male vs. female data points:

```{r}
reactable(sample_theme_options)
```

If you want to change these values, please follow these steps.


Step 1: Create a `.csv` file with two columns (`option` and `value`), modelled after this sample dataset.

> **Important!**: Your .csv must have identical columns and rows as the sample data, or else the plots won't work!

Step 2: Read in the `.csv` file in using `utils::read_csv()`. This will now be an object in your R environment.

Step 3: **Important!!** You must convert the first column (`option`) into the `rownames`. This is a mandatory step to allow the theme_options to be indexed by row name in plotting functions.

Step 4: Run the following code:

```{r, eval=T}
library(tibble)

my_custom_theme_options <- read.csv("my_custom_theme_options.csv") %>% remove_rownames %>% column_to_rownames(var="option")
```

Step 5: Check the resulting object. You should now have 11 objects of 1 variable, and the row names should be `gray_shading_colour`, `line_col`, etc.

Step 6: Go to a plotting function like `plot_raw_current_data()` and replace `sample_theme_options` with your newly created object from Step 5.

I changed `line_col` to orange, and I also changed `male_shape` and `female_shape`. See, the graph looks different now!

```{r}
plot_summary_current_data(
  data = sample_pruned_eEPSC_df$all_cells,
  plot_category = 2,
  plot_treatment = "Control",
  current_type = "eEPSC",
  y_variable = "amplitude",
  hormone_added = "Insulin",
  hormone_or_HFS_start_time = 5,
  include_representative_trace = "no",
  y_axis_limit = 175,
  signif_stars = "yes",
  t_test_df = sample_eEPSC_t_test_df,
  large_axis_text = "no",
  shade_intervals = "no",
  treatment_colour_theme = sample_treatment_names_and_colours,
  theme_options = my_custom_theme_options
)
```



## What does this package replace?

This package replaces manual plot creation and data analysis in GraphPad Prism and/or Excel. Previously, data analysis would have required transferring data between programs, which is time-consuming and potentially error-prone. This package can be combined with report writing in R to generate fully reproducible manuscripts.

```{r, include=F}
knitr::knit_exit()
```


# Package FAQ

## Who's Ruby?

Ruby is a rat, and she's also become an unofficial lab mascot for the Crosby lab! She loves helping out with data analysis, welcoming newcomers to the lab, and keeping lab members company. She found a way to sneak into some of the documentation for this package!


